3.9: Logarithmic Models - R Practice

### 2019-12-03 static/r/21-r-practice-answers.Rmd
Changed name of R practice answers 21 to fit link

---
title: "3.9: Logarithmic Models - R Practice"
author: "YOUR NAME"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    #theme: 
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

---
title: "3.9: Logarithmic Models - R Practice"
author: "Ryan Safner"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    #theme: 
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

@@

```{r}
library(tidyverse)

speed<-read_csv("https://metricsf19.classes.ryansafner.com/data/speeding_tickets.csv")
```

@@

```{r}
reg_linear<-lm(Amount ~ Age, data = speed)
summary(reg_linear)
```

$\widehat{\text{Amount}_i}=131.71-0.29\text{Age}_i$

For every year of age, expected fines decrease by $0.29.

@@

```{r}
# make Age_sq variable
speed<-speed %>%
  mutate(Age_sq=Age^2)

# run quadratic regression
reg_quad<-lm(Amount ~ Age + Age_sq, data = speed) 
summary(reg_quad) 
```

@@

```{r}
reg_quad_alt<-lm(Amount ~ Age + I(Age^2), data = speed) 
summary(reg_quad_alt) 
```

Yes, this is the same output.

@@

$$\widehat{\text{Amount}_i}=146.75-1.17\text{Age}_i+0.01\text{Age}_i^2$$

@@

Yes, a slight improvement. $R^2$ went from 0.004 on the linear model to 0.006 on the quadratic model.

@@

The marginal effect is measured by the first derivative of the regression equation with respect to Age. But you can just remember the resulting formula and plug in the parameters:

$$\begin{align*}
	\frac{d \, Y}{d \, X} &= \beta_1+2\beta_2 X\\
	\frac{d \, Amount}{d \, Age} &= -1.17+2(0.01) \, Age\\
	&=-1.17+0.02 \, Age\\
\end{align*}$$

@@

For 18 year olds:

$$\begin{align*}
	\frac{d \, Amount}{d \, Age} &=-1.17+0.02(18)\\
 		&=-1.17+0.36\\
 		&=-\$0.81\\
\end{align*}$$

For 40 year olds:

$$\begin{align*}
	\frac{d \, Amount}{d \, Age} &=-1.17+0.02(40)\\
 		&=-1.17+0.80\\
 		&=-\$0.37\\
 	\end{align*}$$

```{r}
# Let's do this in R:

# we need broom
library(broom)
tidy_reg_quad<-tidy(reg_quad)

tidy_reg_quad

# save beta 1 
quad_beta_1<-tidy_reg_quad %>%
  filter(term=="Age") %>%
  pull(estimate)

# save beta 2
quad_beta_2<-tidy_reg_quad %>%
  filter(term=="Age_sq") %>%
  pull(estimate)

# create function to estimate marginal effects
marginal_effect<-function(x){
  return(quad_beta_1+2*quad_beta_2*x)
}

# run the function on the 18-year-old and the 40-year-old
marginal_effect(c(18,40))

# close enough, we had some rounding error
```

@@

We can set the derivative equal to 0, or you can just remember the formula and plug in the parameters:

$$\begin{align*}
	\frac{d Y}{d X}  &= \beta_1+2\beta_2 X\\
	0 &=\beta_1+2\beta_2 X\\
	-\beta_1&=2\beta_2 X\\
	-\frac{1}{2} \frac{\beta_1}{\beta_2}&=Age^*\\
	-\frac{1}{2} \frac{-1.17}{0.01} &= Age^*\\
	-\frac{1}{2} 117 & \approx Age^*\\
	58.5 & \approx Age ^*\\
\end{align*}$$

```{r}
# Let's do this in R:

-0.5*(quad_beta_1/quad_beta_2)

# again, some rounding error
```

@@

```{r, fig.retina=3}
ggplot(data=speed)+
  aes(x = Age,
      y = Amount)+
  geom_point()+
  geom_smooth(method="lm",
              formula="y~poly(x,2)",
              color="red")
```

@@

```{r, fig.retina=3}
ggplot(data=speed)+
  aes(x = Age,
      y = Amount)+
  geom_smooth(method="lm",
              formula="y~poly(x,2)",
              color="red")
```

@@

```{r}
reg_cube<-lm(Amount ~ Age + I(Age^2) + I(Age^3), data = speed)
summary(reg_cube)
```

$$\widehat{\text{Amount}_i}=151.95-1.61\text{Age}_i+0.02\text{Age}_i^2-0.00008\text{Age}^3$$

The $t$-statistic on Age$^3$ is small (-1.03) and the $p$-value is 0.31, so the cubic term does not have a significant impact on `Amount`. We should *not* include it.

Just for fun, would the cubic model *look* any better?

```{r, fig.retina=3}
ggplot(data=speed)+
  aes(x = Age,
      y = Amount)+
  geom_point()+
  geom_smooth(method="lm", formula="y~x+I(x^2)", color="red")+
  geom_smooth(method="lm", formula="y~x+I(x^2)+I(x^3)", color="orange")
```

```{r, fig.retina=3}
ggplot(data=speed)+
  aes(x = Age,
      y = Amount)+
  geom_smooth(method="lm", formula="y~x+I(x^2)", color="red")+
  geom_smooth(method="lm", formula="y~x+I(x^2)+I(x^3)", color="orange")
```

@@

```{r}
# install.packages("car") # install once if you don't have
library("car") # load car package
linearHypothesis(reg_cube, c("I(Age^2)", "I(Age^3)")) # F-test 
```

We get a large $F$ of 26.43, with a very small $p$-value. Therefore, we can reject the null hyothesis that the model is linear $(\beta_2=0, \beta_3=0)$. We should in fact *not* use a linear model. Note it does *not* tell us if the model should be *quadratic* or *cubic* (or even *logarithmic* of some sort), *only* that it is not linear. Remember, this was a *joint* hypothesis of all *non-linear* terms $(beta_2$ and $\beta_3)$!

@@

```{r}
# create log of MPHover
speed<-speed %>%
  mutate(log_mph=log(MPHover))

# Run linear-log regression
linear_log_reg<-lm(Amount ~ log_mph, data = speed)
summary(linear_log_reg)

# note we could have done this without creating the variable
# just take log() inside the regression:
linear_log_reg_alt<-lm(Amount ~ log(MPHover), data = speed)
summary(linear_log_reg)
```

$$\widehat{\text{Amount}_i}=-200.10+115.75\text{ln(MPHover}_i)$$

A 1% increase in speed (over the speed limit) increases the fine by $\frac{115.75}{100}\approx \$1.16$

@@

```{r}
# create log of Amount
speed<-speed %>%
  mutate(log_Amount=log(Amount))

# Run log-linear regression
log_linear_reg<-lm(log_Amount ~ MPHover, data = speed)
summary(log_linear_reg)

# again we could have done this without creating the variable
# just take log() inside the regression:
log_linear_reg_alt<-lm(log(Amount) ~ MPHover, data = speed)
summary(log_linear_reg_alt)
```

$$\widehat{\text{ln(Amount}_i)}=3.87+0.05\text{MPHover}_i$$

For every 1 MPH in speed (over the speed limit), expected fine increases by $0.05 \times 100\%=5\%$

@@

```{r}
# Run log-log regression
log_log_reg<-lm(log_Amount ~ log_mph, data = speed)
summary(log_log_reg)

# again we could have done this just taking log()s inside the regression:
log_log_reg_alt<-lm(log(Amount) ~ log(MPHover), data = speed)
summary(log_log_reg_alt)
```

$$\widehat{\text{ln(Amount}_i)}=2.31+0.86\text{ln(MPHover}_i)$$

For every 1% increase in speed (over the speed limit), expected fine increases by 0.86%.

@@

We can compare the $R^2$'s of the three models or compare scatterplots with the regression lines. I will make a table of the three regressions with `huxreg` for easy comparison of fit:

```{r}
library(huxtable)
huxreg("Linear-Log" = linear_log_reg,
       "Log-Linear" = log_linear_reg,
       "Log-Log" = log_log_reg,
       coefs = c("Constant" = "(Intercept)",
                 "MPH Over" = "MPHover",
                 "Log(MPH Over)" = "log_mph"),
       statistics = c("N" = "nobs",
                      "R-Squared" = "r.squared",
                      "SER" = "sigma"),
       number_format = 2)
```

It appears the linear-log model has the best fit with the highest $R^2$ out of the three, but not by very much.

If we wanted to compare scatterplots:

```{r, fig.height=2}
ggplot(data = speed)+
  aes(x = log_mph,
      y = Amount)+
  geom_point()+
  geom_smooth(method="lm", color="red")+
  labs(title="Linear-Log Model")
```

```{r, fig.height=2}
ggplot(data = speed)+
  aes(x = MPHover,
      y = log_Amount)+
  geom_point()+
  geom_smooth(method="lm", color="red")+
  labs(title="Log-Linear Model")
```

```{r, fig.height=2}
ggplot(data = speed)+
  aes(x = log_mph,
      y = log_Amount)+
  geom_point()+
  geom_smooth(method="lm", color="red")+
  labs(title="Log-Log Model")
```

@@

```{r}
full_reg<-lm(Amount~Age+Age_sq+MPHover+Black+Hispanic, data=speed)
summary(full_reg)

library(car)
linearHypothesis(full_reg, c("Black", "Hispanic"))
```

@@

```{r}
# make standardized variables
speed<-speed %>%
  mutate(std_Amount = scale(Amount),
         std_Age = scale(Age),
         std_MPH = scale(MPHover))
```

@@

```{r}
# make standardized variables
std_reg<-lm(std_Amount ~ std_Age + std_MPH, data = speed)
summary(std_reg)
```
